---
title: "Body Fat Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Body Fat Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fedesoriano/body-fat-prediction-dataset\>*

### Reference:

*\<fedesoriano. (2020). Body Fat Prediction Dataset. Retrieved from Kaggle https://www.kaggle.com/datasets/fedesoriano/body-fat-prediction-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load dataset
body_data <- read.csv("bodyfat.csv", colClasses = c(
  Density = "numeric",
  Age = "integer",
  Weight = "numeric",
  Height = "numeric",
  Neck = "numeric",
  Chest = "numeric",
  Abdomen = "numeric",
  Hip = "numeric",
  Thigh = "numeric",
  Knee = "numeric",
  Ankle = "numeric",
  Biceps = "numeric",
  Forearm = "numeric",
  Wrist = "numeric",
  BodyFat = "numeric"
))

# Display the structure of the dataset
str(body_data)

# View the first few rows of the dataset
head(body_data)

# View the dataset in a separate viewer window
View(body_data)
```

## Measures of Central Tendency
```{r MOF}
# Mean for numeric variables
means <- sapply(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")], mean)
means

# Median for numeric variables
medians <- sapply(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")], median)
medians
```

## Measures of Distibution
```{r MOD}
# Standard deviation for numeric variables
sds <- sapply(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")], sd)
sds

# Range for numeric variables
ranges <- sapply(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")], function(x) diff(range(x)))
ranges

# Quartiles for numeric variables
quartiles <- apply(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")], 2, quantile)
quartiles
```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numeric variables
correlation_matrix <- cor(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")])
correlation_matrix

```

## Plots
```{r Plots}
# Scatterplot matrix for numeric variables
pairs(~Density + Age + Weight + Height + Neck + Chest + Abdomen + Hip + Thigh + Knee + Ankle + Biceps + Forearm + Wrist + BodyFat, data = body_data)

# Histogram for a numeric variable (Age)
hist(body_data$Age, main = "Histogram of Age", xlab = "Age")

# Boxplot for a numeric variable (Weight)
boxplot(body_data$Weight, main = "Boxplot of Weight")

# Scatterplot matrix for multiple numeric variables
pairs(~Density + Age + Weight + Height + Neck + Chest + Abdomen + Hip + Thigh + Knee + Ankle + Biceps + Forearm + Wrist + BodyFat, data = body_data)

# Correlation heatmap for numeric variables
library(ggplot2)
# Load the reshape2 package
library(reshape2)
correlation_data <- cor(body_data[, c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist", "BodyFat")])
ggplot(data = melt(correlation_data), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Correlation Heatmap")

# Density plots for multiple numeric variables
library(ggplot2)
library(tidyr)
body_data_long <- gather(body_data, key = "Variable", value = "Value", -Density, -Age)
ggplot(data = body_data_long, aes(x = Value, fill = Variable)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots for Numeric Variables")
```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing values}
# Check for missing values in the dataset
missing_values <- sum(is.na(body_data))

# Output the number of missing values
if (missing_values > 0) {
  print(paste("There are", missing_values, "missing values in the dataset."))
} else {
  print("There are no missing values in the dataset.")
}

```

# Training Model
## Data Splitting
```{r Data Splitting}
# Set seed for reproducibility
set.seed(123)

# Split the dataset into training and testing sets (e.g., 70% training, 30% testing)
library(caTools) # You may need to install this package if not already installed
split <- sample.split(body_data$BodyFat, SplitRatio = 0.7)
training_data <- subset(body_data, split == TRUE)
testing_data <- subset(body_data, split == FALSE)

print(dim(training_data))
print(dim(testing_data))
```

## Bootstrapping
```{r Bootstrapping}
#load required packages
library(boot)
# Set seed for reproducibility
set.seed(123)

# Perform bootstrapping for the entire dataset
bootstrap_results <- boot(body_data, statistic = function(data, i) mean(data[i, "BodyFat"]), R = 1000)
bootstrap_results
```

## Cross-validation
```{r Cross-validation}
# Set seed for reproducibility
set.seed(123)

# Perform k-fold cross-validation (example: linear regression)
library(caret)
control <- trainControl(method = "cv", number = 10)
cv_model <- train(BodyFat ~ ., data = body_data, method = "lm", trControl = control)
summary(cv_model)
print(cv_model)
```

## Training Models
```{r Training Models}
# Load required libraries
library(caret)
library(gbm)
library(randomForest)

# Set seed for reproducibility
set.seed(123)

# Define predictors and response variable
predictors <- c("Density", "Age", "Weight", "Height", "Neck", "Chest", "Abdomen", "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist")
response <- "BodyFat"

# Create training control
train_control <- trainControl(method = "cv", number = 10)

# Train linear regression model (lm)
lm_model <- train(as.formula(paste(response, "~", paste(predictors, collapse = "+"))), 
                  data = body_data, 
                  method = "lm", 
                  trControl = train_control)

# Train gradient boosting model (gbm)
gbm_model <- train(as.formula(paste(response, "~", paste(predictors, collapse = "+"))), 
                   data = body_data, 
                   method = "gbm", 
                   trControl = train_control,
                   verbose = FALSE)

# Train random forest model (rf)
rf_model <- train(as.formula(paste(response, "~", paste(predictors, collapse = "+"))), 
                  data = body_data, 
                  method = "rf", 
                  trControl = train_control)

# Print model summaries
print("Linear Regression Model:")
print(lm_model)
print("Gradient Boosting Model:")
print(gbm_model)
print("Random Forest Model:")
print(rf_model)
```

## Performance Comparison
```{r Performance Comparison}
# Compare model performances using resamples
resamples_list <- resamples(list(LM = lm_model, GBM = gbm_model, RF = rf_model))
summary(resamples_list)
```

## Saving Model
```{r Saving Model}
# Saving the gbm model
saveRDS(gbm_model, file = "./models/gbm_model.rds")

# Load the saved gbm model
loaded_gbm_model <- readRDS("./models/gbm_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Density = 1.0708,
  Age = 23,
  Weight = 154.25,
  Height = 67.75,
  Neck = 36.2,
  Chest = 93.1,
  Abdomen = 85.2,
  Hip = 94.5,
  Thigh = 59,
  Knee = 37.3,
  Ankle = 21.9,
  Biceps = 32,
  Forearm = 27.4,
  Wrist = 17.1
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_gbm_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```


